Day 4 Class notes
--------------------------

```{r include=FALSE}
require(mosaic, quietly=TRUE)
```

### Announcements

* Quiz on Friday.  You will want to have R working by tomorrow so that you can study.
* We'll harvest the first chapter's exercises on Monday.  So make sure that you are able to work with ScoreR.
* Preceptor: We'll be starting organizing the preceptors this week.  They will start having office hours next week.
* Review session by Google Hangout on Tuesday at 9:00-9:30pm.

## Where you should be getting

* Can read in a data frame to R, find variable names, etc.
```{r message=FALSE}
apple = fetchData("M155/Stocks/aapl.csv")
names(apple)
```
* Can calculate simple summary statistics on a single variable: mean, sd, a quantile. etc.
```{r}
mean( Open, data=apple )
sd( Open, data=apple)
```
* Can produce a simple graphic of a distribution, e.g. bwplots, density plots.
* Can break down a calculation or display by the levels of a categorical variable, 
   * `mean(wage~sex, data=CPS85 )`
   * For graphics, there's a "group" argument and the "given" notation.
   * `tally` is weird.  There's nothing to the left of the `~`.  "and" or "plus" do the same thing, "given" does something different.
```{r}
cps = fetchData("CPS85")
tally(~sex, data=CPS85)
tally(~sector, data=CPS85)
tally(~sex&sector, data=CPS85)
tally(~sector&sex, data=CPS85)
tally(~sex|sector, data=CPS85)
tally(~sector|sex, data=CPS85)
```
* Understand why a 95% coverage interval is preferable to a range for many purposes.
* Understand how the standard deviation corresponds to a display and what are the units of the variance.  Understand that the terminology is old-fashioned:
    * "Standard" means "typical".
    * "Deviation" means "distance from a typical value"
    * Nobody uses this term, but the analogous term for a "mean" would be "standard value."
* Know that the variance is just the square of the standard deviation.  For technical reasons, we'll be using the variance and related measures a lot in this course.







### Dealing with Outliers

* If it's just a mistake, mark it as such.  Best to do this in your analysis rather than in your original data.
* Use medians and IQR rather than mean and variance.  This doesn't generalize very well, however, to modeling relationships.
* Transforms, e.g. logarithms.  At the end of the course, we'll introduce a simple and general approach based on ranks.
* Logarithms are widely used in economics, but for a somewhat different reason than dealing with outliers.


### Order statistics

Take everyone out into the hall and line them up from shortest to tallest.  Assign each person a **rank**, which is just their order in the line.  When there is a tie, the order of the people involved in the tie is arbitrary, so average the naive ranks that would be assigned to the people involved in the tie.

Point out the min, max, median, first and third quantile.

#### The Variance and Standard Deviation

Sometimes it's nice to be able to summarize a distribution with just a small set of numbers.  Some possibilities:

* max and min (what's the problem with this?)
* first and third quartile
* median and IQR
* mean and standard deviation

We'll be making extensive use of the mean and standard deviation in this course.  The reason to prefer these won't become apparent until a few weeks into the semester.

"Standard Deviation" -> "Typical Spread" in a more modern terminology.

In French, it's literally "typical spread": [ėcart type](http://fr.wikipedia.org/wiki/%C3%89cart_type)

Give the formulas for mean and standard deviation
$$ m = \frac{1}{n}\sum_{k=1}^{n} x_k$$
$$ v = \frac{1}{n-1} \sum_{k=1}^{n} (x_k - m)^2 $$
$$ s = \sqrt{v} $$

Eyeballing: Standard deviation on a bell-shaped distribution: more or less the half-width at half-height.

Units of m and s.  

Examples of estimation of s.
* Ages of workers in the CPS85 data set.  Compare to max and min.
* Imagine the distribution of speeds of cars on a highway with smooth traffic flow.
* Ages of students at Macalester. (about 1 year)

[In-class activity]((https://dl.dropbox.com/u/5098197/ISM/Numerical-Summaries.pdf) on the properties of the various measures.




### Measurement and measurement bias
* Instrumental or procedural error. Measure heights of people with a ruler, but with the ruler not level.
* Subject or response bias: "How many of you are intravenous drug users?"
* Global warming. Measurement gauges have been encroached on by urbanization.


### Sampling and sampling bias
* Self-selection bias for surveys
* Problems with convenience samples: e.g., interview students at library on Friday night about their study habits
* Selection and survival bias: we don't see the kids who drop out of school, tracking a random sample of Alzheimer's patients to see how long they survive --- you're more likely to sample those who survive a long time.

### Random sampling
* Sampling frame and formal methods.  Use `sample` 

#### In-Class Activity

* Sampling Bias Activity `s = select.books()` which needs to be proceeded by
```{r}
fetchData("simulate.r")
```

[Instructor's write up](../Activities/SelectingLibraryBooks.html)



### What do your fellow students know about statistics
```{r}
ks = fetchData("/Users/kaplan/Dropbox/Stat155Fall2012/knowledge-survey-2012-09-11.csv")
for (item in levels(ks$ProblemItem)) {
  print( item )
  print(tally(~as.character(AnswerContents), data=subset(ks, ProblemItem==item)))
  print("")
}
```


