\documentclass{article}
\pagestyle{empty}
\usepackage[margin=1in]{geometry}

\newcommand{\answerSpace}{\vspace*{1.5cm}}
\begin{document}

\begin{centering}
{\Large \sffamily Math 155 --- Intro. to Statistical Modeling --- Prof. Kaplan}\\
\bigskip

{\Large \sffamily \bfseries{End-of-Semester Course Evaluation: Fall 2013}}\\
\bigskip
\end{centering}

\bigskip

To help improve this course, please answer thoughtfully.  There are two sides to this evaluation.

\begin{enumerate}

\item Do you feel you have a better understanding for statistics and statistical method than when you started the course?

\answerSpace

\item Did anything surprise you positively about the course?

\answerSpace

\item Did anything surprise you negatively?

\answerSpace

\item This course is required for several majors.  If you are following such a major, does having the requirement make sense to you?

\answerSpace

\item Do you think Math 155 will help you in the future?

\answerSpace

\item Are you more interested in studying or using statistics than you were when you enrolled?

\answerSpace

\item How helpful were these tools for you?  (0 means useless, 4 means very helpful.)

\begin{tabular}{lrrrrr}
$\bullet$ Textbook & 0 & 1 & 2 & 3 & 4\\
$\bullet$ Weekly quizzes & 0 & 1 & 2 & 3 & 4\\
$\bullet$ Course website         & 0 & 1 & 2 & 3 & 4\\
$\bullet$ Preceptors         & 0 & 1 & 2 & 3 & 4\\
$\bullet$ In-class activities         & 0 & 1 & 2 & 3 & 4\\
$\bullet$ RStudio & 0 & 1 & 2 & 3 & 4\\
$\bullet$ Homework problems & 0 & 1 & 2 & 3 & 4\\
$\bullet$ Assignment web server & 0 & 1 & 2 & 3 & 4\\
$\bullet$ Professor         & 0 & 1 & 2 & 3 & 4\\
\end{tabular}

\bigskip

\item How did the class size (35 students) affect your experience?

\answerSpace

\item We are considering offering this course in an interactive, online form during J-terms.  Do you think this would be 
attractive or useful for students?

\answerSpace

\item To facilitate updates, we are considering encouraging the publisher to provide the textbook in an e-book form.  This would cost about 
\$20 per copy.  If this had been available this semester, would you have used it?

\answerSpace

\end{enumerate}

Anything else you would like to say about the course?

\end{document}