# Day 34 notes, Nov 22, 2013
```{r include=FALSE}
require(mosaic)
opts_chunk$set(tidy=FALSE)
```


## Review Example  

Difference in age between husband and wife in couples getting married.

Ask: Who is older in a married couple, the man or the woman?  By how much?  Students will think they know the answer.

Let's see if the data support this:
```{r}
m = fetchData("marriage.csv")
mod0 = mm( Age ~ Person, data=m )
mod0
confint(mod0)
mod1 = lm( Age ~ Person, data=m )
summary(mod1)$coef
```
The point estimate is about right, but the margin of error is so large that we can't take this estimate very seriously.  The p-value is so large that we can't reject the null that there is no relationship between `Person` and age.

Try adding in some other variables, astrological sign, years of education, etc. and show that this doesn't help much.

Finally, add in the `BookpageID` variable.  
```{r}
mod2 = lm( Age ~ Person + BookpageID, data=m )
anova(mod2)
head(confint(mod2))
```

This gives an individual ID to each marriage.  Putting this in the model effectively holds the couple constant when considering the `Person`.  In terms of ANOVA, `BookpageID` is eating up lots and lots of variance.

## How Linear Regression can work on Yes/No Variables

### Our Data from Wednesday's class

       | Southern | Northern
-------|:--------:|:-------:
Water  |   4      | 1
Land   |   2      | 1

Here's a spreadsheet: [link to edit](https://docs.google.com/spreadsheet/ccc?key=0Am13enSalO74dG1SM19taXVadVp0QkloM3BQNl9LZnc&usp=sharing), [URL to load](https://docs.google.com/spreadsheet/pub?key=0Am13enSalO74dG1SM19taXVadVp0QkloM3BQNl9LZnc&single=true&gid=0&output=csv) using `fetchGoogle()`.

```{r}
globe <- fetchGoogle("https://docs.google.com/spreadsheet/pub?key=0Am13enSalO74dG1SM19taXVadVp0QkloM3BQNl9LZnc&single=true&gid=0&output=csv")
summary(lm(Water=="W" ~ 1, data=globe))
```
Compare to the obvious answer: 5 out of 8 are water.

Notice that the confidence interval covers the conventional wisdom of 2/3. 

### Exploring Some Hypotheses

#### Hypothesis I

The Northern hemisphere and the Southern hemisphere have different fractions of the Earth's surface.  In other words, the area of the Southern hemisphere is different from 50% of total area.  (Silly, but so might be many scientific hypotheses.  Try it out.)

```{r}
summary(lm(Hemisphere=="S" ~ 1, data=globe))$coef
```

* What's the Null corresponding to the given p-value?
* How to calculate an appropriate p-value for conventional wisdom?
* How much data should we collect to have a fair chance of rejecting the Null?  What's the alternative hypothesis?  Calculate sample size to make the standard error small enough to rule out the Null in favor of the Alternative.



#### Hypothesis II

Do we have evidence to support the hypothesis that the Southern hemisphere area is not 50% of the Earth's surface?

```{r}
summary(lm(Water=="W" ~ 1 + Hemisphere, data=globe))$coef
```

How much data should we collect to be able to reject the Null?  What's your Alternative.




