# Statistical Geometry and Computing

```{r include=FALSE}
require(mosaic)
vectorAngle <- function(v1,v2,degrees=FALSE,subtractMean=FALSE) {
  if (subtractMean) {
    v1 <- v1 - mean(v1)
    v2 <- v2 - mean(v2)
  }
  L1 <- sqrt(sum(v1^2))
  L2 <- sqrt(sum(v2^2))
  cosTheta <- sum(v1*v2)/(L1*L2)
  ang <- acos(cosTheta)
  if (degrees) return(180*ang/pi)
  else return(ang)
  }
```

### Instructor's Guide

In this activity, students use the computer to confirm fundamental relationships in statistical geometry.

### Objectives

1. Demonstrate that the residual vector and the fitted model values vector are at right angles, regardless of the model chosen.  This means that the fitting process itself generates the right-angle relationship.  
2. See that R^2 is an angle between two vectors.

## Activity

### Preliminaries

Explain how to measure an angle.  You can do it with a protractor.  When computing the angle between two vectors, this only works if you can somehow get a protractor in to the space between the two vectors.  For two- and three-space, this is easy.  But for four and higher space, what could it mean.

Instead, we use the dot product.  The rule is $\cos(\theta) = \frac{v_1 \cdot v_2}{| v_1 | | v_2 |}$.

For students who don't know about the dot product, explain the operation algebraically and then demonstrate it on some coordinate pairs for which we do and don't know the angle.



Start by having the students read in some data of interest and compute some model that makes sense to them.  For example:
```{r}
houses = fetchData("SaratogaHouses.csv")
mod = lm( Price ~ Living.Area*Baths + Bedrooms*Fireplace, data=houses)
```
The angle between the fitted and residuals
```{r}
vectorAngle( fitted(mod), resid(mod), degrees=TRUE)
```



Sums of squares add up in the pythagorean way
```{r}
sum( fitted(mod)^2 ) + sum( resid(mod)^2 )
with(sum( Price^2 ), data=houses)
```
R^2 relates to the cosine of the angle between the fitted model values and the response variable --- once the mean is subtracted out
```{r}
r.squared(mod)
ang <- vectorAngle(with(data=houses,Price),fitted(mod), 
                   subtractMean=TRUE)
R <- cos(ang)
R^2
```
