# Counting Model Vectors

```{r include=FALSE}
require(mosaic)
```

We're going to pull out a small subset of the `CPS85` data in order to explore model vectors and their relationship to model terms.

First, to construct the `small` data set, follow these instructions **exactly**.

```{r}
set.seed(1234)
small <- droplevels(sample(CPS85, size=10))
small <- transform(small, ageGrp=as.character(ntiles(age,3)))
rownames(small) <- NULL
```

In this small sample, you can see how many levels there are of each categorical variable with the command `with( small, levels( variableName ) )`

```{r}
with( small, levels( sector ) )
with( small, levels( union ) )
```

For each of the following models, your job is 
* **FIRST**, to predict how many explanatory model vectors there will be in the model for this `small` data set.
* Second, confirm or refute your prediction by using the `model.matrix()` command.
* Third, explain any discrepancy between your prediction and the results of `model.matrix()`.
* Fourth, find the standard deviation of the residuals.  If the model is a "perfect" fit, explain why.

### EXAMPLE:

How many explanatory model vectors in the model `wage ~ 1`?

* Prediction: There's just one, the intercept.  A vector of all 1s.

* Confirming
```{r}
mod = lm( wage~1, data=small )
model.matrix(mod)
```

The prediction was right.

The standard deviation of residuals is
```{r}
sd( resid(mod) )
```
Not at all a perfect fit.  Indeed, the model "explains" nothing, since the residuals have a standard deviation that is every bit as large as the response variable itself:
```{r}
sd( wage, data=small )
```

### Model 1: `wage ~ sex`

### Model 2: `wage ~ sector`

### Model 3: `wage ~ union`

### Model 4: `wage ~ ageGrp`

### Model 5: `wage ~ educ + sex`

### Model 6: `wage ~ age + ageGrp`

### Model 7: `wage ~ educ * sex`

### Model 8: `wage ~ age * sector`

### Model 9: `wage ~ age * sector + age * sex`

### Model 10: `wage ~ ageGrp*sector + ageGrp*sex`

### Finally ...

Try fitting the last model to `CPS85` (rather than `small`) and look at the standard deviation of the residuals.  Comment on the difference from the residuals fitted to `small`).