# Day 33 Notes, Nov. 20, 2013

```{r include=FALSE}
require(mosaic)
opts_chunk$set(tidy=FALSE,dev="svg",out.width="60%")
```


## Review of the three main reports

### Regression Report

Gives coefficient and standard error for individual coefficients.

Null Hypothesis: The coefficient being examined would be zero with a large enough sample.  The other vectors are doing what they will.

t distribution: 
* Chi by eye: bigger in magnitude than 2 suggests significance.
* Parameter: degrees of freedom of residual.

Shape:
```{r}
plotFun(dt(x, 1) ~ x, x.lim=c(-10,10))
plotFun(dt(x, 3) ~ x, add=TRUE, col="red")
plotFun(dt(x, 10) ~ x, add=TRUE, col="green", lwd=2)
plotFun(dnorm(x, mean=0, sd=1) ~ x, add=TRUE, col="black", lwd=3)
```

#### Calculation of p:

1. Divide coef by standard error --- that's the t.
2. Evaluate the t distribuion at the observed t statistic at **both** tails.

```{r}
mod = lm( width ~ length + sex + domhand + biggerfoot, 
          data=KidsFeet )
summary(mod)$coef
# How many cases
nrow(KidsFeet)
# How many vectors
sum(!is.na(coef(mod)))
# Number of residual vectors : 39-5 => 34
2*pt(-abs(4.702993), df=34)
```
Report says: 4.155301e-05

Confidence intervals apply here: use shuffle.

Where does t come from?

```{r}
small = sample(KidsFeet, 8)
mod0 = lm( width ~ length + sex + domhand + biggerfoot, 
           data=KidsFeet )
trials = do(500)*(summary(lm( shuffle(width) ~ length + sex + shuffle(domhand) + biggerfoot, 
    data=small))$coef[4,3])
histogram(~result, data=trials)
plotFun(dt(x,3)~x, add=TRUE, col="red")
plotFun(dnorm(x)~x, add=TRUE, col="blue")
```

Implement the Null.  Extract t. Repeat many times. 

Plot the distribution of t values.  

### R^2

The model considered as a whole.  Translate to F for ease of finding p-values.

Null Hypothesis: None of the vectors is any better than junk.

Could get a confidence interval with resampling.

### ANOVA Report

Duplicating it by hand.

Partial R^2

Null Hypothesis: The group of vectors being examined is no different from junk vectors.  We estimate the size of junk vectors by considering the size of residuals.

#### Calculation of p:

Evaluate the F distribution with the appropriate degrees of freedom in the numerator and denominator at the observed F statistic at **one tail**, the high one, of the distribution.

Order dependence: race and wages in CPS85

```{r}
mod0 = lm( wage ~ race + sex + sector + educ + exper, data=CPS85 )
summary(mod0)
anova(mod0)
mod1 = lm( wage ~ sex + sector + educ + exper + race, data=CPS85 )
summary(mod1)
anova(mod1)
anova(mod1)$"F value"
```

## Categorical Responses

Yes/No.  Conversion to 1/0.

`lm()` gives the proportion of 1s.

Toss globe to find proportion of Earth covered by water.

Separate tallies for Northern and Southern hemisphere.

### Our Data

       | Southern | Northern
-------|----------|---------
Water  |   4      | 1
Land   |   2      | 1


Example: Whickham: model by smoking.  Then by smoking and age.  Make a prediction from the model fitted by lm().

Show logistic regression

```{r}
m1 <- lm( outcome=="Alive" ~ smoker + age, data=Whickham )
f1 <- makeFun(m1)
f1("Yes", 50)
plotFun( f1("Yes", age) ~ age, age.lim=range(20,100))
m2 <- glm( outcome=="Alive" ~ smoker + age, data=Whickham, family="binomial" )
f2 <- makeFun(m2)
plotFun( f2("Yes", age) ~ age, age.lim=range(20,100))
```

