# Day 29, Nov. 11, 2013

```{r include=FALSE}
require(mosaic)
```

## Review of Vocabulary

* Two hypotheses:
   * Null Hypothesis.  The star of the show.
        Some ways to think about it:
        * nothing is going on
        * coefficients are zero
        * vectors are random
        You either **reject** or **fail to reject** the Null.
    * The Alternative Hypothesis.  A supporting actor.
        * Your best guess of what's happening
        * The smallest interesting effect
        * The results of a preliminary study.
        You use this to design your study.  The Alternative needs to be very specific and drawn from expert knowledge.  There is a tradition in statistics education of referring to the alternative as an "anything but" kind of hypothesis.  This is a bad idea.
* Two kinds of error
    * Type I: Reject the Null even though the Null is right
    * Type II: Fail to reject the Null even though the Alternative is right.
* p-value: probability of the observed value or stronger under the Null
    * significance level --- the p-value is the "achieved significance level".  Comes from the idea of using confidence intervals to do a hypothesis test: related to the confidence level.
* power: probability of rejecting the Null under the Alternative Hypothesis.

## The Summary Report and the p-value

Show the t-statistic and the translation into a p-value.

Show the p-value reported on R^2.

Notice that the alternative plays no role whatsoever in the regression report.

### App for playing with Significance and Power

```{r eval=FALSE}
fetchData("mHypTest.R")
mHypTest() # by default, a coefficient
```

### Shuffling 

A proof for the existence of Extra-Sensory Perception!  If I can get you to focus on a number, I can predict, to some extent, your thought process.

Your birthday is a number that plays an important part in your thought process.  Generate a random number between 0 and your birday.

[Permutation test by hand](http://dl.dropbox.com/u/5098197/ISM/Activities/permutation-test.pdf)

Spreadsheet reading command:
```{r}
ESPdata <- fetchGoogle("https://docs.google.com/spreadsheet/pub?key=0Am13enSalO74dGJPam1QSkdaSlNlVVQ1Ni1HVHJpNlE&single=true&gid=0&output=csv")
```

### Power

How did I know that I could reject the Null in the shuffling problem?  I did a little simulation.
```{r}
mysim <- function(n=15) {
   days = resample(1:31,size=n)
   nums = ceiling( runif(n,min=0,max=days))
   mod = lm(nums~days)
   list(r2=r.squared(mod), p=summary(mod)$coef[2,4])
}
s15 = do(1000)*mysim(30) # typical R^2 is about 0.4
mean(~r2, data=s15)
tally( ~ p<0.05, data=s15,format='proportion')
```

Do `mHypTest(TRUE)` setting the "effect size" to about 0.4

### R^2 as a test statistic

Activity: fit the kids' feet data with different numbers of model terms: `rand(p)` 

Find an average R^2 for each p and draw the graph of R^2 versus p.

### From R^2 to F




Construction of F

```{r}
n <- nrow(KidsFeet)
degreesOfFreedom=5
trials = do(5000)* lm(width ~ rand(degreesOfFreedom), data=KidsFeet)
F = with(data=trials,
         (r.squared/degreesOfFreedom)/((1-r.squared)/(n - (degreesOfFreedom+1))))
histogram(F)
plotFun( df(x,5,32) ~ x, add=TRUE, col="red", lwd=2)

```

Comparing the distribution of F to the distribution of R^2.  F is much more stable, easier to interpret.  We look for F near 1 if the null is true.

